---
title: "Comparison of ML Classifiers Using Sparklyr"
output: 
  html_notebook:
    toc: true
    toc_depth: 2
    toc_float:
      collapsed: false
---

# Overview

You can use `sparklyr` to fit a wide variety of machine learning algorithms in Apache Spark. This analysis compares the performance of six classification models in Apache Spark on the [Titanic](https://www.kaggle.com/c/titanic) data set.

For the Titanic data, decision trees and random forests performed the best and had comparatively fast run times. See [results](#compare-results) for a detailed comparison.

ID | Function | Description | AUC Rank | Run time Rank
---|---------|-------------|----------|-------------
1 | Random forest | [ml_random_forest](http://spark.rstudio.com/reference/sparklyr/latest/ml_random_forest.html) | 1 | 3
2 | Decision tree | [ml_decision_tree](http://spark.rstudio.com/reference/sparklyr/latest/ml_decision_tree.html) | 2 | 2
3 | Gradient boosted tree | [ml_gradient_boosted_trees](http://spark.rstudio.com/reference/sparklyr/latest/ml_gradient_boosted_trees.html) | 3 | 6 
4 | Logistic regression | [ml_logistic_regression](http://spark.rstudio.com/reference/sparklyr/latest/ml_logistic_regression.html) | 4 | 4
5 | Multilayer perceptron (neural net) | [ml_multilayer_perceptron](http://spark.rstudio.com/reference/sparklyr/latest/ml_multilayer_perceptron.html) | 5 | 5
6 | Naive Bayes | [ml_naive_bayes](http://spark.rstudio.com/reference/sparklyr/latest/ml_naive_bayes.html) | 6 | 1



```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
#knitr::opts_chunk$set(eval = FALSE)
```

```{r parquet, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# Convert titanic_train data into parquet format and output to disk
library(sparklyr)
library(titanic)
parquet_path <- "./titanic-parquet"
parquet_table <- "titanic"
if(!dir.exists(parquet_path)){
  sc <- spark_connect(master = "local", version = "2.0.0")
  copy_to(sc, titanic_train, parquet_table, overwrite = TRUE)
  tbl(sc, parquet_table) %>% spark_write_parquet(path = parquet_path)
}
```


***

# Load the data

Load the popular Titanic data set into a local spark cluster. 

First, download the [titanic](https://cran.r-project.org/web/packages/titanic/index.html) package from CRAN. These data can be used to predict survival based on factors including: class, gender, age, and family. A thorough background is available on [Kaggle](https://www.kaggle.com/c/titanic). 

Second, create local Spark cluster. This example uses Apache Spark 2.0.0 which includes advanced feature transforms and methods which will be used later in the analysis.

Third, load the data into the local spark context. Data can be loaded from CSV or [Parquet](https://parquet.apache.org/) format.

```{r data, message=FALSE, warning=FALSE}
library(sparklyr)
library(dplyr)
library(tidyr)
library(titanic)
library(ggplot2)
library(purrr)

# Connect to local spark cluster and load data
#sc <- spark_connect(master = "local", version = "2.0.0")
#spark_read_parquet(sc, name = "titanic", path = "titanic-parquet")
#titanic_tbl <- tbl(sc, "titanic")
titanic_tbl <- titanic_train
```

***

# Tidy the data

Tidy the data in preparation for model fitting. `sparkyr` uses `dplyr` syntax when connecting to the Spark SQL API and specific functions functions for connecting to the Spark ML API. 

## Spark SQL transforms

Use feature transforms with Spark SQL. Create new features and modify existing features with `dplyr` syntax.

ID  | Feature | Action
----|---------|-----------
1 | Family_Size | Create number of siblings and parents
2 | Pclass | Format passenger class as character not numeric
3 | Embarked | Remove a small number of missing records
4 | Age | Impute missing age with average age

```{r sparkSQL}
# Transform features with Spark SQL API
titanic2_tbl <- titanic_tbl %>% 
  mutate(Family_Size = SibSp + Parch + 1L) %>% 
  mutate(Pclass = as.character(Pclass)) %>%
  filter(!is.na(Embarked)) %>%
  mutate(Age = if_else(is.na(Age), mean(Age, na.rm = TRUE), Age)) #%>%
  #sdf_register("titanic2")
```

> Tip: `sdf_register` is used to save our table for later analysis.


## Spark ML transforms

Use feature transforms with Spark ML. Use `ft_bucketizer` to bucket family sizes into groups.

```{r sparkFT}
# Transform family size with Spark ML API
titanic_final_tbl <- titanic2_tbl %>%
  #mutate(Family_Size = as.numeric(Family_size)) %>%
  mutate(
    Family_Sizes = cut(Family_Size, c(0,1,2,5,12))
    ) %>%
  mutate(Family_Sizes = as.character(as.integer(Family_Sizes))) #%>%
  #sdf_register("titanic_final")
```

> Tip: You can use magrittr pipes to chain dplyr commands with sparklyr commands. For example, `mutate` is a dplyr command that accesses the Spark SQL API whereas `sdf_mutate` is a sparklyr command that accesses the Spark ML API.

## Train-validation split

Randomly partition the data into train and test sets.

```{r partition}
# Partition the data
partition <- titanic_final_tbl %>% 
  #mutate(Survived = as.numeric(Survived), SibSp = as.numeric(SibSp), Parch = as.numeric(Parch)) %>%
  select(Survived, Pclass, Sex, Age, SibSp, Parch, Fare, Embarked, Family_Sizes) %>%
  #sdf_partition(train = 0.75, test = 0.25, seed = 8585)
  mutate(Survived = as.factor(Survived),
         Sex = as.factor(Sex),
         Embarked = as.factor(Embarked),
         Family_Sizes = as.factor(Family_Sizes),
         Pclass = as.factor(Pclass))

set.seed(8585)
ind <- sample(nrow(partition), nrow(partition) * 0.75, replace = FALSE)

# Create table references
train_tbl <- partition[ind,]
test_tbl <- partition[-ind,]
```

> Tip: Use `sdf_partition` to create training and testing splits.

***

# Train the models

Train multiple machine learning algorithms on the training data. Score the test data with the fitted models.

## Logistic regression

Logistic regression is one of the most common classifiers. Train the logistic regression and examine the predictors.

```{r train}
# Model survival as a function of several predictors
ml_formula <- formula(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked + Family_Sizes)

# Train a logistic regression model
(ml_log <- glm(ml_formula, train_tbl, family='binomial'))
#predict.glm(ml_log, test_tbl, type = 'response')
```

## Other ML algorithms

Run the same formula using the other machine learning algorithms. Notice that training times vary greatly between methods.

```{r ml}
library(rpart)
library(randomForest)
library(gbm)
library(e1071)
library(neuralnet)

## Decision Tree
ml_dt <- rpart(ml_formula, train_tbl)
predict(ml_dt, test_tbl)[,2]

## Random Forest
ml_rf <- randomForest(ml_formula, train_tbl)
predict(ml_rf, test_tbl, type = 'prob')[,2]

## Gradient Boosted Tree
#ml_gbt <- gbm(ml_formula, train_tbl, distribution = 'bernoulli', n.trees = 1000)
#predict(ml_gbt, test_tbl, 1000)

## Naive Bayes
ml_nb <- naiveBayes(ml_formula, train_tbl)
predict(ml_nb, test_tbl, type = 'raw')[,2]

## Neural Network
#ml_nn <- neuralnet(ml_formula, train_tbl)

```

## Validation data

Score the test data with the trained models.

```{r score}
# Bundle the modelss into a single list object
ml_models <- list(
  "Logistic" = ml_log,
  "Decision Tree" = ml_dt,
  "Random Forest" = ml_rf,
  #"Gradient Boosted Trees" = ml_gbt,
  "Naive Bayes" = ml_nb#,
  #"Neural Net" = ml_nn
)

# Create a function for scoring
#score_test_data <- function(model, data=test_tbl){
  #pred <- predict(model, data)
  #select(pred, Survived, prediction)
#}

# Score all the models
#ml_score <- lapply(ml_models, score_test_data)
?factor
ml_score <- as.tbl(data.frame(
  "Survived" = as.numeric(levels(test_tbl$Survived))[test_tbl$Survived],
  "Logistic" = predict.glm(ml_log, test_tbl, type = 'response'),
  "Decision Tree" = predict(ml_dt, test_tbl)[,2],
  "Random Forest" = predict(ml_rf, test_tbl, type = 'prob')[,2],
  "Naive Bayes" = predict(ml_nb, test_tbl, type = 'raw')[,2]
))

```

***

# Compare results

Compare the model results. Examine performance metrics: lift, AUC, and accuracy. Also examine feature importance to see what features are most predictive of survival.

## Model lift

Lift compares how well the model predicts survival compared to random guessing. Use the function below to estimate model lift for each scored decile in the test data. The lift chart suggests that the tree models (random forest, gradient boosted trees, or the decision tree) will provide the best prediction.


```{r lift}
# Lift function

calculate_lift <- function(data, model){
  data %>%
    mutate_(x = model) %>%
    mutate(bin = ntile(desc(x), 10)) %>% 
    group_by(bin) %>% 
    summarize(count = sum(Survived)) %>% 
    mutate(prop = count / sum(count)) %>% 
    arrange(bin) %>% 
    mutate(prop = cumsum(prop)) %>% 
    select(-count) %>% 
    collect() %>% 
    as.data.frame()
}

# Initialize results
ml_gains <- data.frame(bin = 1:10, prop = seq(0, 1, len = 10), model = "Base")

# Calculate lift
for(i in names(ml_score)[-1]){
  ml_gains <- calculate_lift(ml_score, i) %>%
    mutate(model = i) %>%
    rbind(ml_gains, .)
}

# Plot results
ggplot(ml_gains, aes(x = bin, y = prop, colour = model)) +
  geom_point() + geom_line() +
  ggtitle("Lift Chart for Predicting Survival - Test Data Set") + 
  xlab("") + ylab("")
```

> Tip: `dplyr` and `sparklyr` both support windows functions, including `ntiles` and `cumsum`.


# Discuss

You can use `sparklyr` to run a variety of classifiers in Apache Spark. For the Titanic data, the best performing models were tree based models. Gradient boosted trees was one of the best models, but also had a much longer average run time than the other models. Random forests and decision trees both had good performance and fast run times.

While these models were run on a tiny data set in a local spark cluster, these methods will scale for analysis on data in a distributed Apache Spark cluster.