---
title: "Analysis of flights data using Apache Spark"
output: html_notebook
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, eval=TRUE)
```

## Motivation

Is there evidence to suggest that some airline carriers make up time in flight? This analysis predicts time gained in flight by airline carrier. The flights data consume 12 gigabytes of uncompressed data and represent 123 million flights over 22 year period.

## Connect to spark

Set the environment variable `SPARK_HOME` and then run `spark_connect`. After connecting you will be able to browse the Hive metadata in the RStudio Server Spark pane.

```{r}
# Connect to Spark
library(sparklyr)
library(dplyr)
library(ggplot2)
Sys.setenv(SPARK_HOME="/usr/lib/spark")
config <- spark_config()
sc <- spark_connect(master = "yarn-client", config = config, version = '1.6.2')
```

## Cache the tables into memory

Use `tbl_cache` to load the flights table into memory. Create a dplyr reference to the Spark DataFrame.

```{r}
# Cache flights Hive table into Spark
tbl_cache(sc, 'flights')
flights_tbl <- tbl(sc, 'flights')

# Cache airlines Hive table into Spark
tbl_cache(sc, 'airlines')
airlines_tbl <- tbl(sc, 'airlines')

# Cache airports Hive table into Spark
tbl_cache(sc, 'airports')
airports_tbl <- tbl(sc, 'airports')
```

## Create a model data set

Filter the data to contain only the records to be used in the fitted model. Join carrier descriptions for reference. Create a new variable called `gain` which represents the amount of time gained (or lost) in flight.

```{r}
# Filter records and create target variable 'gain'
model_data <- flights_tbl %>%
  filter(!is.na(arrdelay) & !is.na(depdelay) & !is.na(distance)) %>%
  filter(depdelay > 15 & depdelay < 240) %>%
  filter(arrdelay > -60 & arrdelay < 360) %>%
  filter(year >= 2003 & year <= 2007) %>%
  left_join(airlines_tbl, by = c("uniquecarrier" = "code")) %>%
  mutate(gain = depdelay - arrdelay) %>%
  select(year, month, arrdelay, depdelay, distance, uniquecarrier, description, gain)

# Summarize data by carrier
model_data %>%
  group_by(uniquecarrier) %>%
  summarize(description = min(description), gain=mean(gain), 
            distance=mean(distance), depdelay=mean(depdelay)) %>%
  select(description, gain, distance, depdelay) %>%
  arrange(gain)
```

## Train a linear model

Predict time gained or lost in flight as a function of distance, departure delay, and airline carrier.

```{r}
# Partition the data into training and validation sets
model_partition <- model_data %>% 
  sdf_partition(train = 0.8, valid = 0.2, seed = 5555)

# Fit a linear model
ml1 <- model_partition$train %>%
  ml_linear_regression(gain ~ distance + depdelay + uniquecarrier)

# Summarize the linear model
summary(ml1)
```

## Assess model performance

Compare the model performance using the validation data.

```{r}
# Calculate average gains by predicted decile
model_deciles <- lapply(model_partition, function(x) {
  sdf_predict(ml1, x) %>%
    mutate(decile = ntile(desc(prediction), 10)) %>%
    group_by(decile) %>%
    summarize(gain = mean(gain)) %>%
    select(decile, gain) %>%
    collect()
})

# Create a summary dataset for plotting
deciles <- rbind(
  data.frame(data = 'train', model_deciles$train),
  data.frame(data = 'valid', model_deciles$valid),
  make.row.names = FALSE
)

# Plot average gains by predicted decile
deciles %>%
  ggplot(aes(factor(decile), gain, fill = data)) +
  geom_bar(stat = 'identity', position = 'dodge') +
  labs(title = 'Average gain by predicted decile', x = 'Decile', y = 'Minutes')
```

### Visualize the best and worst gains by airline carrier

Compare actual gains to predicted gains for an out of time sample.

```{r}
# Select 2008 as test data
data_2008 <- flights_tbl %>%
  filter(!is.na(arrdelay) & !is.na(depdelay) & !is.na(distance)) %>%
  filter(depdelay > 15 & depdelay < 240) %>%
  filter(arrdelay > -60 & arrdelay < 360) %>%
  filter(year == 2008) %>%
mutate(uniquecarrier = ifelse(crsarrtime == 351, "DH", uniquecarrier)) %>%
mutate(uniquecarrier = ifelse(crsarrtime == 120, "HP", uniquecarrier)) %>%
mutate(uniquecarrier = ifelse(crsarrtime == 347, "TZ", uniquecarrier)) %>%
  left_join(airlines_tbl, by = c("uniquecarrier" = "code")) %>%
  mutate(gain = depdelay - arrdelay) %>%
  select(origin, dest, year, month, arrdelay, depdelay, distance, 
         uniquecarrier, description, gain)

# Score the 2008 data
score_2008 <- sdf_predict(ml1, data_2008)

# Summarize data by carrier
plot_2008 <- score_2008 %>%
  group_by(description) %>%
  summarize(gain = mean(gain), prediction = mean(prediction), freq = n()) %>%
  filter(freq > 10000) %>%
  collect

# Plot actual gains and predicted gains by airline carrier
ggplot(plot_2008, aes(gain, prediction)) + 
  geom_point(alpha = 0.75, color = 'red', shape = 3) +
  geom_abline(intercept = 0, slope = 1, alpha = 0.15, color = 'blue') +
  geom_text(aes(label = substr(description, 1, 20)), size = 3, alpha = 0.75, vjust = -1) +
  labs(title='Average Gains Forecast', x = 'Actual', y = 'Predicted')
```

## Conclusions

Some carriers make up more time than others in flight, but the differences are relatively small. The average time gains between the best and worst airlines is only six minutes. The best predictor of time gained is not carrier but flight distance. The biggest gains were associated with the longest flights.

## Build rollups for flex dashboard

```{r}
# Summarize by origin, destination, and carrier
summary_2008 <- score_2008 %>%
  rename(carrier = uniquecarrier, airline = description) %>%
  group_by(origin, dest, carrier, airline) %>%
  summarize(
    flights = n(),
    distance = mean(distance),
    avg_dep_delay = mean(depdelay),
    avg_arr_delay = mean(arrdelay),
    avg_gain = mean(gain),
    pred_gain = mean(prediction)
    )

# Materialize Spark DataFrame as a Hive table
sdf_save_table(summary_2008, "summary_2008", overwrite = TRUE)

# Collect and save objects
pred_data <- collect(summary_2008)
airports <- collect(select(airports_tbl, name, faa, lat, lon))
ml1_summary <- capture.output(summary(ml1))
save(pred_data, airports, ml1_summary, file = 'flights_pred_2008.RData')
```


